# Data

We used a publicly available [dataset](www.retentionprediction.org/hplc/database/) that comprises the measurements of RP-HPLC retention times collected for 1026 analytes. The retention times were measured under isocratic conditions on Eclipse Plus C18 (Agilent) stationary phase with 3.5 μm particles. The experiments were conducted using a mixture of two solvents: solvent A, which was made of 0.1% formic acid in water, and solvent B, which was made of 0.1% formic acid in acetonitrile. The column temperature was set at 35\^{\circ}C. The data were collected by Boswell et al. and were used to create a method to predict retention time by Back-Calculating the Gradient.

```{r message=FALSE, warning=FALSE}
# load packages
library(here)
library(ggplot2)

# load data
DS        <- read.csv(here::here("1_data/database_stan_1026.csv"),header = TRUE, sep = ";", dec = ".")
DS_names  <- read.csv(here::here("1_data/database_stan_1026_analyte_names.csv"),header = TRUE, sep = ",", dec = ".")

# divide the analytes into groups according to MM
DS$MM_group <- with(DS, ifelse(MW_ACD < 200, "1",
                                 ifelse(MW_ACD < 300 & MW_ACD >= 200, "2", 
                                        ifelse(MW_ACD < 400 & MW_ACD >= 300, "3", "4" ))))
lab <- function(x){
  ifelse(x == 1, "MM < 200",
         ifelse(x == 2, "200 \u2264 MM < 300", 
                ifelse(x == 3, "300 \u2264 MM < 400", "400 \u2264 MM" )))
}

nObs <- length(DS$ID)
nAnalytes <- length(unique(DS$ID))
fi <- seq(0,1,0.1)

ggplot(data=DS, aes(x=concentration, y=logk, group=ID)) + 
  geom_line(aes(color=ID)) + 
  facet_wrap(~MM_group, ncol=2, labeller=labeller(MM_group = lab)) +
  labs(x = expression(varphi), y = expression(Log~k[Obs])) + 
  theme_gray(base_size = 14) + theme(legend.position="none") 
```

The pH value of the mobile phase was verified experimentally for the purpose of this work. It equaled 2.66 with a standard deviation of 0.19 for the range of acetonitrile contents from 5 to 95%.

```{r message=FALSE, warning=FALSE}
DS_pH <- read.csv(here::here("1_data/pH.csv"),header = TRUE, sep = ";", dec = ",")


ggplot(data=DS_pH, aes(x=fi, y=pH)) + labs(x = expression(varphi), y = "pH") + 
  geom_point() + geom_hline(yintercept=mean(DS_pH$pH), color = "red") +
  geom_hline(yintercept=mean(DS_pH$pH)-sd(DS_pH$pH), linetype="dashed", color = "red") +
  geom_hline(yintercept=mean(DS_pH$pH)+sd(DS_pH$pH), linetype="dashed", color = "red") + ylim(c(2.0,3.5))
```

The values of molecular mass MM were added to the dataset. They were calculated using [ACD/Labs program](www.acdlabs.com) based on the provided structures of analytes. The MM ranged from 73.09 to 656.8.

```{r message=FALSE, warning=FALSE}
library(tidyverse)

df <- DS %>% distinct(ID,MW_ACD) %>%
  gather(MW_ACD, key= ACD, value="value") %>%
  mutate(ACD = recode(ACD, "MW_ACD" = "MM"))

ggplot(data= df , aes(x=ACD, y=value))+
  geom_boxplot(aes(1))+facet_wrap(~ ACD,scales = "free") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
Information on the functional groups of individual compounds has also been added to the data.

```{r message=FALSE, warning=FALSE}
functional_groups         <- read.csv(here::here("1_data/checkmol_nr_groups.csv"),header = TRUE, sep = ",", dec = ".")
functional_groups_names   <- read.delim(here::here("1_data/checkmol_functional_group_names.csv"),header = TRUE, sep = ";", dec = ".")
```

This information was first prepared for analysis in several steps:

1) combine nr of caroboxylic acid and carboxyalic acid salt functional groups
```{r message=FALSE, warning=FALSE}
functional_groups[,76]=functional_groups[,76]+functional_groups[,77]       
functional_groups[which(functional_groups[,202]>8.1),202] = 8; # heterocyclic compounds with more than 8 heterocycles are treated as if they have eigth
```

2) exclude functional groups that repeat itself (some groups are nested)
```{r message=FALSE, warning=FALSE}
idx_excluded <- c(1, 2, 3, 6, 27, 28, 37, 47, 48, 51, 55, 61, 62, 67, 73, 74, 75, 77, 80, 91, 99, 109, 116, 117, 121, 125, 129, 142, 153, 154, 160, 161, 168, 173, 178, 181, 182, 186, 187, 191, 196)
functional_groups_names[which(functional_groups_names$ID_group %in% idx_excluded),2]

functional_groups <- functional_groups[,-idx_excluded]
functional_groups_names <- functional_groups_names[-idx_excluded,]
```

3) exclude functional groups not present on any analyte from the dataset

```{r message=FALSE, warning=FALSE}
idx_not_present = which(colSums(functional_groups==0) == nrow(functional_groups))
functional_groups_names[which(functional_groups_names$ID_group %in% idx_not_present),2]

functional_groups <- functional_groups[,-idx_not_present]
functional_groups_names <- functional_groups_names[-idx_not_present,]
```
The functional groups present in the analyzed chemical compounds are presented below, along with their frequencies

```{r message=FALSE, warning=FALSE}
sumdata=data.frame(value=apply(functional_groups,2,sum))
sumdata$key=functional_groups_names[,2]
sumdata <- sumdata[order(sumdata$value),]
sumdata$split <- rep(c(1,2),each=50)

layout(c(1,2), widths=c(1,3))

p1 <- ggplot(data=sumdata[which(sumdata$split==1),], aes(x=reorder(key, +value), y=value, fill=key)) +
  geom_bar(fill="blue", stat="identity") + 
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank()) + 
  coord_flip()+
  scale_y_continuous(breaks=c(0,5,10))

p2 <- ggplot(data=sumdata[which(sumdata$split==2),], aes(x=reorder(key, +value), y=value, fill=key)) +
  geom_bar(fill="blue", stat="identity") + 
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank()) + 
  coord_flip() + scale_y_continuous(trans='log10',breaks=c(0,10,100,1000))

library(patchwork)

# Custom design (# means empty area)
design <- "
  1222
"

# Combine the plots with a custom layout
p1 + p2 +
  plot_layout(design = design)
```
